---
title: "Michelson - 1880"
author: "Daniel"
date: 2018-01-11
output: 
  github_document:
    toc: true
---

```{r setup, include = FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)

```

```{r load packages and set parameters, message = FALSE, warning = FALSE, echo = FALSE}
# Libraries
library(tidyverse)

# Parameters
  # File with Michelson 1880 speed of light data
file_data <- "../../data/michelson-1880/data.rds"
```

In the summer of 1879, Albert A. Michelson performed 100 experiments to estimate
the speed of light. He published his results in 1880 in [Experimental
Determination of the Velocity of Light](https://books.google.com/books?id=343nAAAAMAAJ&lpg=PA115&pg=PA109#v=onepage&q&f=false).
Michelson was 26 years old at the time of the experiments and went on to win the
Nobel prize in Physics in 1907.

This challenge is to examine Michelson's data.

__q1__ First read in the data into a tibble `df` using the `file_data` parameter above.

```{r load data} 
df <- as_tibble(readRDS(file_data))
```

Each of his 100 experiments resulted in an estimate for the speed of light in
air in kilometers per second (kps). Since his goal was to estimate the speed of
light in a vacuum, which is faster than in air, he added a correction of 80 kps.
He also added a correction of 12 kps to account for the effects of temperature
on his instruments.

__q2__ For your analysis, add a new variable `kps_vacuum` to `df` with the above
corrections to obtain Michelson's estimates for the speed of light in a vacuum.

```{r speed corrections}
df <- df %>%
  mutate(kps_vacuum = kps_air + 80 + 12)
```

## The data

__q3__ Look at the data and get an overview of it. (Hint: Useful functions
include `glimpse()` and `summary()`.) Are there any problems with the data?

```{r}
glimpse(df)
summary(df)
```

It looks like we have a date for each experiment (`date`), whether it was run in 
the morning or afternoon/night (`time`), the temperature that day (`temp`), the 
speed of the light in air (`kps_air`), and the corrected speed in a vacuum 
(`kps_vacuum`).

One small problem with the data is that the time (AM/PM") is missing for the first experiment. I'm not a physicist, so I'm not sure if this is a problem with the dataset, but Michelson did not take an equal number of measurments on each day he ran the experiment. Furthermore, there is an uneven distribution of times (AM/PM) across dates such that on some dates all the measurments were taken during PM and others during AM. In fact, there are no PM measurments at the the tail ends of the experiment. Also, all of these measurments are from early June to early July, so maybe time of year affects the measurments


__q4__ Are Michelson's measurements of kps_vacuum consistent over time?

```{r time, warnings = F}
df %>%
  ggplot(mapping = aes(x = time, y = kps_vacuum)) +
  geom_boxplot()
```

If the question is whether the values of the `kps_vacuum` variable are consistent over the `time` variable, it appears that Michelson's measurments of kps_vacuum are slower and more variable in the morning than they are in the afternoon/evening. If human error can be introduced to these measurments, maybe he is slower and less alert in the monring than in the afternoon/evening. It could also be the case that temperature is lower in the morning and so the speed of light is slower in the monring.

```{r}
model1 <- lm(temp ~ time, data = df)
summary(model1)
```

It appears that this is the case - temperature is higher in the evening.

However, we can also look at whether the temperatures are consistent across date and time. We can fit lines to the measurments over time...

```{r date, warning = FALSE, message = FALSE}
df %>%
  ggplot(mapping = aes(x = date, y = kps_vacuum, color = time)) +
  geom_point() +
  geom_smooth()
```

...or we can use boxplots.

```{r}
df %>%
  ggplot(mapping = aes(x = factor(date), y = kps_vacuum, color = time)) +
  geom_boxplot() +
  coord_flip() + 
  labs(x = "date", y = "speed")
```

We can see that Michelson took only PM measurements for the first and last few days of his experiment. This graph also shows us that the PM measurments are faster on average and have lower error than AM measurments. Furthermore, we can see that the PM measurments decrease in speed over the course of the experiment (from beginning June to beginning July). The AM measurments vary across the experiment, but seem to stay the same on average.


__q5__ Does the temperature affect the measurements?

```{r temperature, warnings = FALSE, message = FALSE}
df %>%
  ggplot(mapping = aes(x = temp, y = kps_vacuum)) +
  geom_point() + 
  geom_smooth()
```

Temperature does appear to have an effect on the measurments though it is not enormous and is rather flat for the mid-range of temperatures. It looks like that speed of light increases as time increases. We can even run a linear regression to see if this effect is statistically significant.

```{r temp regression}
model1 <- lm(kps_vacuum ~ temp, data = df)
summary(model1)
```

The effect of temperature on the measurments is statistically significant such that high temperatures lead to faster measurements. This correction that we made for the effect of temperature on instrument seems to be problematic in light of this effect. We made a fixed correction of 12 kps across all measurments, but different temperatures could affect the instruments differently.


__q6__ Plot a distribution of kps_vacuum. How much variability is there around
the mean value?

```{r mean kps, warning = F, message = F}
mean_kps_vacuum <- mean(df$kps_vacuum)

df %>% 
  ggplot(mapping = aes(x = kps_vacuum)) +
  geom_histogram(binwidht = 1) + 
  geom_vline(xintercept = mean_kps_vacuum)
```

There is a fair amount of variability around the mean. It's not the case that all the measurments are the same. We can calculate the standard deviation to quantify the precise amount of variability.

```{r std dev kps vacuum}
sd(df$kps_vacuum)
```

A standard deviation of 80 kilometers per second seems substantial, but comparing it to the scale of the measurments, which are around 300,000 makes it seem pretty small.


## Compare with true value for the speed of light in a vacuum

__q7__ Look up the true value for the speed of light in a vacuum. Find the mean
value of Michelson's estimates of kps_vacuum. He estimated his error to be ± 51
kps. Is the true value within his error?

The true speed of light in a vacuum is 299792.458 kps. Michelson's mean was 299944.4 kps. The difference between these two values is:

```{r difference, echo = FALSE}
diff <- 299792.458 - 299944.4
```

The true speed of light in a vacuum is `r diff` kps slower than Michelson's mean, so the true vale was not within his error. Maybe this is due to the fact that he took all his measurments in the dead of summer and we showed that temperature has a positive relationship with speed of light.


__q8__ Plot the distribution again, this time including lines for Michelson's
estimate and the true value. Include a legend with the colors for the two lines.

```{r, warning = FALSE, message = FALSE}
vline <- data.frame(
  estimate = c("michelson's estimate", "true speed"),
  val = c(299944.4, 299792.458)
)

true_speed <- 299792.458

df %>% 
  ggplot(mapping = aes(x = kps_vacuum)) +
  geom_bar(position = "dodge") + 
  geom_vline(mapping = aes(xintercept = val, color = estimate), data = vline) 
```

We could also take a look at whether the distributions vary according to `time`.

```{r, warning = FALSE}
df %>% 
  ggplot(mapping = aes(x = kps_vacuum, fill = time)) +
  geom_bar(position = "dodge") 
```

It appears that there is a peak for morning and a peak for evening measurements and that the morning measurment peak is closer to the true mean.
