---
title: "Cholera map"
author: "Daniel"
date: 2018-02-14
output: 
  html_document:
    toc: true
    toc_depth: 6
---

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(compare)
library(leaflet)
if (!require(leaflet.extras)) {
  devtools::install_github('bhaskarvk/leaflet.extras')
  library(leaflet.extras)
}
library(sf)

# Parameters
  # URL for death and pump data in KML format
url_data <- "http://rtwilson.com/downloads/SnowGIS_KML.zip"
  # File for answers
file_answers <- "../../data/cholera-map/answers.rds"

cholera_deaths_kml <- "C:/Users/djole/Google Drive/classes/dcl/c17/SnowGIS_KML/cholera_deaths.kml"
pumps_kml <- "C:/Users/djole/Google Drive/classes/dcl/c17/SnowGIS_KML/pumps.kml"

deathIcon <- makeIcon(
  iconUrl = "C:/Users/djole/Google Drive/classes/dcl/c17/death.png",
  iconWidth = 20, iconHeight = 25,
  iconAnchorX = 22, iconAnchorY = 94
)

pumpIcon <- makeIcon(
  iconUrl = "C:/Users/djole/Google Drive/classes/dcl/c17/pump.png",
  iconWidth = 20, iconHeight = 25,
  iconAnchorX = 22, iconAnchorY = 94
)

#===============================================================================
# Read in answers
answers <- read_rds(file_answers)
```

Maeva Finker (DCL 2017 Winter) contributed to this challenge.

## The 1854 London cholera epidemic

In late August 1854, a severe cholera epidemic broke out in the Soho neighborhood of London. As recounted by Dr. John Snow

> The most terrible outbreak of cholera which ever occurred in this kingdom, is
> probably that which took place in Broad Street, Golden Square, and adjoining
> streets, a few weeks ago. Within two hundred and fifty yards of the spot
> where Cambridge Street joins Broad Street, there were upwards of five hundred
> fatal attacks of cholera in ten days. The mortality in this limited area
> probably equals any that was ever caused in this country, even by the plague;
> and it was much more sudden, as the greater number of cases terminated in a
> few hours. The mortality would undoubtedly have been much greater had it not
> been for the flight of the population. [^1]

How cholera spread was not understood at the time, and the prevailing belief was that it was spread through the air. With the deaths from this outbreak concentrated into a relatively small area, Snow suspected that the cause may have been contaminated water from a community water pump at Broad and Cambridge streets. He collected data to test this theory.

> On proceeding to the spot, I found that nearly all of the deaths had taken
> place within a short distance of the pump. There were only ten deaths in
> houses situated decidedly nearer to another street pump. In five of these
> cases the families of the deceased persons informed me that they always sent
> to the pump in Broad Street, as they preferred the water to that of the pump
> which was nearer. In three other cases, the deceased were children who went to
> school near the pump in Broad Street. Two of them were known to drink the
> water; and the parents of the third think it probable that it did so.
> [^2]

After Snow reported his findings to the authorities responsible for the community water supply, they ordered the handle from the Broad Street pump removed, and the epidemic soon ended.

Snow created a visualization of his data in the form of a map of the deaths from this outbreak and the water pumps in the area. This map helped to overturn the prevailing beliefs about the transmission of cholera. As a result, London improved its water infrastructure and developed procedures that soon eliminated such cholera outbreaks.

## Download and read in death and pump data

Using a browser and the URL in `url_data` in the parameters, download data for the number of deaths, their location, and the location of the pumps. Unzip and save the data in a place other than a repo. The data is in KML format.

### Read in death data

__q1.1__ Use `sf::read_sf()` to read the downloaded file `cholera_deaths.kml` into the sf object `deaths`, and convert it to have the variables:

* deaths
* geometry

Sort `deaths` in order of decreasing deaths.

```{r, message=FALSE}
deaths <- 
  sf::read_sf(cholera_deaths_kml) %>%
  mutate(deaths = as.integer(str_extract(Description, "[0-9]+"))) %>%
  select(deaths, geometry) %>%
  arrange(desc(deaths))

# Compare result with answer
if (exists("deaths")) compare(answers$deaths, deaths)
```

### Read in pump data

__q1.2__ Use `sf::read_sf()` to read the downloaded file `pumps.kml` into the sf object `pumps`, and convert it to have the variable:

* geometry

```{r, message=FALSE}
pumps <-
  sf::read_sf(pumps_kml) %>%
  select(geometry)

# Compare result with answer
if (exists("pumps")) compare(answers$pumps, pumps)
```

## Leaflet

We will use the leaflet package to create our maps. This package calls the Leaflet JavaScript library and for this reason the output type for this document is html_document. You can view HTML documents from any browser, but GitHub will only display the HTML source. To submit this task, upload both the challenge.Rmd and the knit challenge.html.

To understand the basics of using Leaflet, read the first three sections of [Leaflet for R](https://rstudio.github.io/leaflet/map_widget.html). The examples use the older sp format. We explain below how to use the functions with the sf format.

## Plot pumps and deaths

__q2__ Use leaflet to visualize the deaths and pump data. Choose a better base map than the default for displaying this data.

Notes

* With sf data, the `lng` and `lat` arguments are not necessary in `addCircles()`.
* The way you map a variable to an aesthetic in leaflet is with `~`, for example `radius = ~deaths`.
* You can add tooltips by using the `label` argument to `addCircles()`.

Radius is center of most deaths!
```{r}
aoe <- 
  deaths %>%
  slice(1)



leaflet(data = deaths) %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addMarkers(clusterOptions = markerClusterOptions(), icon = deathIcon) %>%
  addMarkers(icon = pumpIcon, data = pumps) %>%
    addCircles(
   data = aoe,
   radius = ~sqrt(deaths) * 60,
   stroke = 0.3,
   color = "orange",
   fillOpacity = 0.05
   )

leaflet(data = deaths) %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircles(radius = ~deaths, color = "red", fillColor = "red") %>%
  addMarkers(icon = pumpIcon, data = pumps, label = "pump") %>%
  addCircles(
   data = aoe,
   radius = ~sqrt(deaths) * 60,
   stroke = 0.3,
   color = "orange",
   fillOpacity = 0.05
  )

```


## Plot pumps and heatmap of deaths

__q3__ Now create a heatmap of the deaths using `addHeatmap()` from the leaflet.extras package.

Notes

* With sf data, the `lng` and `lat` arguments are not necessary in `addHeatmap()`.    


```{r}
leaflet(data = deaths) %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addHeatmap(radius = 12, blur = 15, max = 2) %>% 
  #addCircles(radius = 1, fillColor = "blue", opacity = 1, fill = TRUE) %>%
  addMarkers(icon = pumpIcon, data = pumps, label = "pump") %>%
  addCircles(
   data = aoe,
   radius = ~sqrt(deaths) * 60,
   stroke = 0.3,
   color = "orange",
   fillOpacity = 0.05
  )

```

[^1]: Snow J. [On the Mode of Communication of
Cholera](http://johnsnow.matrix.msu.edu/work.php?id=15-78-52).
London: Churchill; 1855. p. 38.

[^2]: Snow J. [On the Mode of Communication of
Cholera](http://johnsnow.matrix.msu.edu/work.php?id=15-78-52).
London: Churchill; 1855. p. 39-40.

