---
title: "Movie ratings - Part 1"
author: "Your Name"
date: 2018-
output: 
  github_document:
    toc: true
    toc_depth: 6
---

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
options(knitr.table.format = "markdown")
```

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(compare)
library(knitr)

# Parameters
  # Data files for movies, ratings, and reviewers
file_movie <- "../../data/movie-ratings/data/movie.rds"
file_rating <- "../../data/movie-ratings/data/rating.rds"
file_reviewer <- "../../data/movie-ratings/data/reviewer.rds"
  # File for answers
file_answers <- "../../data/movie-ratings/answers_1.rds"

#===============================================================================

# Read in data
movie <- read_rds(file_movie)
rating <- read_rds(file_rating)
reviewer <- read_rds(file_reviewer)

# Read in answers
answers <- read_rds(file_answers)
```

For this challenge you will wrangle data from three toy datasets about movies,
movie ratings, and reviewers. The data and the questions were adapted from the
[SQL Movie-Rating Query
Exercises](https://lagunita.stanford.edu/courses/DB/SQL/SelfPaced/courseware/ch-sql/seq-exercise-sql_movie_query_core/),
which are a part of the Stanford Online course Databases: DB5 SQL.

## Data

### Movies

```{r}
movie %>% kable()
```

### Ratings

```{r}
rating %>% kable()
```

### Reviewers

```{r}
reviewer %>% kable()
```

## Directions

For each question below, store your result in a variable with the same name as
the question. For example, for question q1 store your result in the variable
`q1`. This challenge comes with answers. Each chunk has code to compare your
result with the answer. If your result matches the answer, it will return
`TRUE`.

## Questions

q1 Find the titles of all movies directed by Steven Spielberg.

```{r q1}
q1 <-
  movie %>%
  filter(director == "Steven Spielberg") %>%
  select(title)

# Print results
if (exists("q1")) q1 %>% kable()

# Compare result with answer
if (exists("q1")) compare(answers$q1, q1)
```


q2 Find all years that have a movie that received a rating of 4 or 5, and sort
them in increasing order.

```{r q2}
q2 <-
  movie %>%
  left_join(rating, by = "movie_id") %>%
  filter(stars > 3) %>%
  distinct(year) %>%
  select(year) %>%
  arrange(year)

# Print results
if (exists("q2")) q2 %>% kable()

# Compare result with answer
if (exists("q2")) compare(answers$q2, q2)
```

q3 Find the titles of all movies that have no ratings.

```{r q3}
q3 <-
  movie %>%
  left_join(rating, by = "movie_id") %>%
  filter(is.na(stars)) %>%
  select(title)


# Print results
if (exists("q3")) q3 %>% kable()

# Compare result with answer
if (exists("q3")) compare(answers$q3, q3)
```

q4 Some reviewers didn't provide a date with their rating. Find the names of
all reviewers who have ratings with a NA value for the date.

```{r q4}
q4 <-
  rating %>%
  left_join(reviewer, by = "reviewer_id") %>% 
  filter(is.na(rating_date)) %>%
  select(name) %>%
  arrange(name)

# Print results
if (exists("q4")) q4 %>% kable()

# Compare result with answer
if (exists("q4")) compare(answers$q4, q4)
```

q5 Write a query to return the ratings data in a more readable format:
reviewer name, movie title, stars, and rating_date. Also, sort the data, first
by reviewer name, then by movie title, and lastly by number of stars.

```{r q5}
q5 <- 
  rating %>% 
  full_join(movie, by = "movie_id") %>%
  full_join(reviewer, by = "reviewer_id") %>%
  select(name, title, stars, rating_date) %>%
  arrange(name, title, stars)

# Print results
if (exists("q5")) q5 %>% kable()

# Compare result with answer
if (exists("q5")) identical(answers$q5, q5)
```

q6 For all cases where the same reviewer rated the same movie twice and gave
it a higher rating the second time, return the reviewer's name and the title
of the movie.

```{r q6}
q6 <-
  rating %>%
  left_join(reviewer, by = "reviewer_id") %>%
  group_by(reviewer_id, movie_id) %>%
  count() %>%
  filter(n > 1) %>% 
  left_join(movie, by = "movie_id") %>%
  left_join(rating, by = c("movie_id", "reviewer_id")) %>%
  ungroup() %>%
  select(reviewer_id, title, rating_date, stars) %>%
  group_by(reviewer_id) %>%
  mutate(rating_date = if_else(rating_date == min(rating_date), 
                               "rating_1", "rating_2")) %>%
  spread(rating_date, stars) %>%
  filter(rating_2 > rating_1) %>%
  select(reviewer_id, title) %>%
  left_join(reviewer, by = "reviewer_id") %>%
  ungroup() %>%
  select(name, title)

# Print results
if (exists("q6")) q6 %>% kable()

# Compare result with answer
if (exists("q6")) identical(answers$q6, q6)
```

q7 For each movie that has at least one rating, return the movie title and the
highest number of stars that movie received. Sort by movie title.

```{r q7}
q7 <- 
  movie %>%
  left_join(rating, by = "movie_id") %>%
  filter(!is.na(stars)) %>%
  group_by(title) %>%
  top_n(1, stars) %>%
  select(title, stars) %>%
  distinct(title, stars) %>%
  arrange(title) %>%
  mutate(stars = as.double(stars)) %>%
  ungroup()

# Print results
if (exists("q7")) q7 %>% kable()

# Compare result with answer
if (exists("q7")) identical(answers$q7, q7)
```

q8 For each movie, return the title and the "rating spread", that is, the
difference between highest and lowest ratings given to that movie. Sort by
rating spread from highest to lowest, then by movie title.

```{r q8}
q8 <-
  movie %>%
  left_join(rating, by = "movie_id") %>%
  group_by(title) %>%
  mutate(rating_spread = max(stars) - min(stars)) %>%
  select(title, rating_spread) %>%
  distinct(title, rating_spread) %>%
  filter(!is.na(rating_spread)) %>%
  arrange(desc(rating_spread), title) %>%
  ungroup()

# Print results
if (exists("q8")) q8 %>% kable()

# Compare result with answer
if (exists("q8")) identical(answers$q8, q8)
```

q9 Find the difference between the average rating of movies released before
1980 and the average rating of movies released in 1980 and later. (Make sure to
calculate the average rating for each movie, then the average of those
averages for movies before 1980 and movies after. Don't just calculate the
overall average rating before and after 1980.)

```{r q9}
q9 <-
  movie %>%
  left_join(rating, by = "movie_id") %>%
  group_by(title) %>%
  mutate(mov_avg = mean(stars, na.rm = T)) %>%
  filter(!is.na(mov_avg)) %>%
  distinct(title, year, mov_avg) %>%
  ungroup() %>%
  group_by(year > 1980) %>%
  mutate(year_avg = mean(mov_avg)) %>%
  select(bef_1980 = "year > 1980", year_avg) %>%
  distinct(year_avg) %>%
  spread(bef_1980, year_avg) %>%
  select(a = "FALSE", b = "TRUE") %>%
  transmute(mean_rating_difference = a - b)
  

# Print results
if (exists("q9")) q9 %>% kable()

# Compare result with answer
if (exists("q9")) compare(answers$q9, q9)
```


